Supplementary material to `Adaptive group-regularized logistic elastic net regression'

### 2019-06-28 docs/biostatistics_main.Rnw
commit before second revision

\documentclass[a4paper,hidelinks]{article}
\usepackage{url,bm,algorithm,algpseudocode,bbm,amsmath,amssymb,amsthm,
amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,
titlesec,multicol,verbatim,threeparttable,booktabs,dsfont,parskip,
xr-hyper,hyperref,enumerate}
\externaldocument[sm-]{supplement}

% objects
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\balpha}{\bm{\alpha}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\newcommand{\ones}{\mathbf{1}}

% functions and operators
\renewcommand{\L}{\mathcal{L}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\argmax}{\text{argmax} \,}
\newcommand{\expit}{\text{expit}}
\newcommand{\erfc}{\text{erfc}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\tr}{^{\text{T}}}
\newcommand{\diag}{\text{diag}}
\newcommand{\KL}{D_{\text{KL}}}
\newcommand{\trace}{\text{tr}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

% distributions
\newcommand{\N}{\text{N}}
\newcommand{\TG}{\text{TG}}
\newcommand{\Bi}{\text{Binom}}
\newcommand{\PG}{\text{PG}}
\newcommand{\Mu}{\text{Multi}}
\newcommand{\GIG}{\text{GIG}}
\newcommand{\IGauss}{\text{IGauss}}
\newcommand{\Un}{\text{U}}

% syntax and readability
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}
\renewcommand{\[}{\left[}
\renewcommand{\]}{\right]}
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}} 
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

% settings
\graphicspath{{../figs/}}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Adaptive group-regularized logistic elastic net regression}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Carel F.W. Peeters$^{1}$, 
Aad W. van der Vaart$^{2}$, and \\ Mark A. van de Wiel$^{1,3}$}

@@

	\noindent
	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health
	research institute, Amsterdam University medical centers, PO Box 7057, 1007 MB
	Amsterdam, The Netherlands \\*
	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\*
	3. MRC Biostatistics Unit, Cambridge University, Cambridge, United Kingdom

	\begin{abstract}
		{In high-dimensional data settings, additional information on the features 
		is often available. Examples of such external information in omics research
		are: (a) $p$-values from a previous study, (b) a summary of prior 
		information, and (c) omics annotation. The inclusion of this information in
		the analysis may enhance classification performance and feature selection,
		but is not straightforward in the standard regression setting. As a solution
		to this problem, we propose a group-regularized (logistic) elastic net
		regression method, where each penalty parameter corresponds to a group of
		features based on the external information. The method, termed 
		\texttt{gren}, makes use of the Bayesian formulation of logistic elastic net
		regression to estimate both the model and penalty parameters in an 
		approximate empirical-variational Bayes framework. Simulations and
		applications in two different cancer studies show
		that, if the partitioning of the features is informative, classification 
		performance and feature
		selection are indeed enhanced.}
	\end{abstract}

	\noindent\textbf{Keywords}: Empirical Bayes; High-dimensional data; 
	Prediction; Variational Bayes

	\noindent\textbf{Software available from}: 
	\url{https://CRAN.R-project.org/package=gren}

@@

	Assume we have a partitioning of the features in $G$ groups, such that each
	feature belongs to one group. Let $\G(g)$ be the feature index set of group 
	$g$ for $g=1, \dots, G$ and let $w_g \in \mathbb{R}_{>0}$ denote a 
	group-specific weight. In a generalised elastic net regression, the penalised 
	likelihood is maximised to yield parameter estimates:
	\begin{align}
	  \hat{\bbeta} & := \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) - 
	  \frac{\lambda_1}{2} \sum_{g=1}^G \sum_{j \in \G(g)} |w_g \cdot \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \sum_{j \in \G(g)}\( w_g \cdot \beta_j\)^2 
	  \label{eq:weightedlik}\\
	  & = \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) -  \frac{\lambda_1}{2} 
	  \sum_{g=1}^G \sqrt{\lambda'_g} \sum_{j \in \G(g)} | \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in \G(g)} \beta_j^2, 
	  \nonumber
	\end{align}
	where $\L(\y ; \bbeta)$ denotes the likelihood function of $\bbeta$ given the 
	observed data $\y = \begin{bmatrix} 
	y_1 & \cdots & y_n \end{bmatrix} \tr$, and 
	$\lambda_1, \lambda_2 \in \mathbb{R}_{>0}$ are the `global' penalty 
	parameters. We wrote $w^2_g = \lambda'_g$ to emphasise that these 
	group-specific weights may be interpreted as penalty multipliers. Note that
	the regular elastic net is recovered by setting $\forall g: \lambda'_g=1$.

@@

  The estimated penalty multipliers are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group and the non-expressed group. 
  \texttt{gren} seems to outperform the other methods for a range of smaller 
  model sizes. Selection of larger models is impaired by the large
  $\alpha$, a property inherent to $L_1$-norm penalization. With a smaller
  $\alpha$ larger models are possible (see SM Section \ref{sm-sec:colorectal}). 
  The performance of \texttt{cMCP} is somewhat unstable. This unstable 
  estimation is an issue in all investigated group lasso extensions. 
  Overall, the inclusion of the extra information on the 
  features benefits predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit only slightly for \texttt{GRridge}.
<<barplot_lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Estimated (a) penalty multipliers and (b) AUC in the colorectal cancer example.", out.width="100%", fig.asp=1/2>>=
@

@@

  The estimated penalty multipliers follow the expected pattern: 
  larger cis-correlations receive smaller penalties and smaller $p$-values are
  penalized less. The AUC of
  \texttt{gren} is slighty better than the other methods for a range of model
  sizes. In this example, ridge and \texttt{GRridge} perform almost 
  identical, while \texttt{gren} outperforms the regular elastic net.
<<barplot_lines_rnaseq_oral_cancer_res1_auc, fig.cap="Estimated (a) penalty multipliers for the cis-correlations and (b) external $p$-values, and (c) estimated AUC in the oral cancer example.", out.width="100%", fig.asp=1>>=
@

@@

	\bibliographystyle{author_short3.bst}
	\bibliography{refs}

	\section*{Session info}

<<r session-info, eval=TRUE, echo=TRUE, tidy=TRUE>>=
devtools::session_info()
@

>>>>

\documentclass[oupdraft]{bio}
\usepackage{xr-hyper}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, 
% citecolor=citecolor]{hyperref}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable,
enumerate}

\externaldocument[sm-]{biostatistics_supp}

% Add history information for the article if required
%\history{Received August 1, 2010;
%	revised October 1, 2010;
%	accepted for publication November 1, 2010}

% objects
\newcommand{\x}{\mathbf{x}}
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\m}{\mathbf{m}}

% functions and operators
\newcommand{\tr}{^{\text{T}}}
\newcommand{\expit}{\text{expit}}
\newcommand{\argmax}{\text{argmax} \,}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\diag}{\text{diag}}
\newcommand{\trace}{\text{tr}}

% syntax and readability
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}
\renewcommand{\[}{\left[}
\renewcommand{\]}{\right]}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

@@

	% Add a footnote for the corresponding author if one has been
	% identified in the author list
	\footnotetext{To whom correspondence should be addressed.}

	\begin{abstract}
		{In high-dimensional data settings, additional information on the features 
		is often available. Examples of such external information in omics research
		are: (a) $p$-values from a previous study, (b) a summary of prior 
		information, and (c) omics annotation. The inclusion of this information in
		the analysis may enhance classification performance and feature selection,
		but is not straightforward in the standard regression setting. As a solution
		to this problem, we propose a group-regularized (logistic) elastic net
		regression method, where each penalty parameter corresponds to a group of
		features based on the external information. The method, termed 
		\texttt{gren}, makes use of the Bayesian formulation of logistic elastic net
		regression to estimate both the model and penalty parameters in an 
		approximate empirical-variational Bayes framework. Simulations and
		applications in two different molecular marker and cancer type studies show
		that, if the partitioning of the features is informative, classification 
		performance and feature
		selection are indeed enhanced.}
		{Empirical Bayes; High-dimensional data; Prediction; Variational Bayes}
	\end{abstract}

@@

	Assume we have a partitioning of the features in $G$ groups, such that each
	feature belongs to one group. Let $\G(g)$ be the feature index set of group 
	$g$ for $g=1, \dots, G$ and let $w_g \in \mathbb{R}_{>0}$ denote a 
	group-specific weight. In a generalised elastic net regression, the penalised 
	likelihood is maximised to yield parameter estimates:
	\begin{align}
	  \hat{\bbeta} & := \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) - 
	  \frac{\lambda_1}{2} \sum_{g=1}^G \sum_{j \in \G(g)} |w_g \cdot \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \sum_{j \in \G(g)}\( w_g \cdot \beta_j\)^2 
	  \label{eq:weightedlik}\\
	  & = \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) -  \frac{\lambda_1}{2} 
	  \sum_{g=1}^G \sqrt{\lambda'_g} \sum_{j \in \G(g)} | \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in \G(g)} \beta_j^2, 
	  \nonumber
	\end{align}
	where $\L(\y ; \bbeta)$ denotes the likelihood function of the 
	observed data $\y = \begin{bmatrix} 
	y_1 & \cdots & y_n \end{bmatrix} \tr$, and 
	$\lambda_1, \lambda_2 \in \mathbb{R}_{>0}$ are the `global' penalty 
	parameters. We wrote $w^2_g = \lambda'_g$ to emphasise that these 
	group-specific weights may be interpreted as penalty multipliers. Note that
	the regular elastic net is recovered by setting $\forall g: \lambda'_g=1$.

@@

  The estimated penalty multipliers are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group and the non-expressed group. 
  \texttt{gren} seems to outperform the other methods for a range of smaller 
  model sizes. Selection of larger models is impaired by the large
  $\alpha$, a property inherent to $L_1$-norm penalization. With a smaller
  $\alpha$ larger models are possible (see SM Section \ref{sm-sec:colorectal}). 
  The performance of \texttt{cMCP} is somewhat unstable. This unstable 
  estimation is an issue in all investigated group lasso extensions. 
  Overall, the inclusion of the extra information on the 
  features benefits predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit only slightly for \texttt{GRridge}.

@@

  The estimated penalty multipliers follow the expected pattern: 
  larger cis-correlations receive smaller penalties and smaller $p$-values are
  penalized less. The AUC of
  \texttt{gren} is slighty better than the other methods for a range of model
  sizes. In this example, ridge and \texttt{GRridge} perform almost 
  identical, while \texttt{gren} outperforms the regular elastic net.

@@

	\bibliographystyle{biorefs}
	\bibliography{refs}

<<barplot_lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Estimated (a) penalty multipliers and (b) AUC in the colorectal cancer example.", out.width="100%", fig.asp=1/2>>=
@

<<barplot_lines_rnaseq_oral_cancer_res1_auc, fig.cap="Estimated (a) penalty multipliers for the cis-correlations and (b) external $p$-values, and (c) estimated AUC in the oral cancer example.", out.width="100%", fig.asp=1>>=
@

### 2019-06-28 docs/supplement.Rnw
commit before second revision

<<lines_metabolomics_alzheimer_res1_auc, fig.cap="Estimated AUC for the alzheimer example.", out.width="100%", fig.asp=1>>=
@

<<lines_metabolomics_alzheimer_res1_briers, fig.cap="Estimated Brier skill score for the alzheimer example.", out.width="100%", fig.asp=1>>=
@

>>>>

<<lines_metabolomics_alzheimer_res1_auc, fig.cap="Estimated AUC for the alzheimer example with with (a) $\\alpha=0.05$, (b) $\\alpha=0.5$, and (c) $\\alpha=0.95$.", out.width="100%", fig.asp=1>>=
@

<<lines_metabolomics_alzheimer_res1_briers, fig.cap="Estimated Brier skill score for the alzheimer example with (a) $\\alpha=0.05$, (b) $\\alpha=0.5$, and (c) $\\alpha=0.95$.", out.width="100%", fig.asp=1>>=
@
